---
title: "Assignment 1 Task 2"
author: "Annika Leiby"
date: "1/25/2021"
output: html_document
---

### Data Source

Data for this task comes from Google Earth Engine and can be accessed at https://www.kaggle.com/zanderventer/environmental-variables-for-world-countries/data.


This task will complete a Principal Component Analysis (PCA), create a biplot displaying results of the PCA, interpret the PCA and biplot findings, and create a scree plot to visualize variance among the Principal Components (PC). The environmental variables that will be analyzed are elevation, slope, tree canopy cover, annual mean rainfall, annual mean temperature, and wind.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)


# Change echo to = FALSE, and add message = FALSE and warning = FALSE so prevent messages coming up in knitted document 
```


```{r}
# Attach Packages 

library(tidyverse)
library(dplyr)
library(janitor)
library(here)
library(ggfortify)
library(gghighlight)
library(lubridate)
library(plotly)

```

```{r}
# Read in the data 
# Use select() function to select variables you want to run a PCA on. I selected for country, elevation, slope, tree_canopy_cover, rain_mean_annual, temp_mean_annual,and wind. 
# Use drop_na() function to remove rows with N/As.

env_country <- read_csv("world_env_vars.csv") %>%
  clean_names() %>%
  select(country, elevation, slope, tree_canopy_cover, rain_mean_annual, temp_mean_annual, wind) %>%
    drop_na() 
```

```{r}
# Use select() and set to -country because it is categorical and should not be included in the PCA.
# Use scale() function to get all the variable units on the same scale ready for PCA. 
# Use function prcomp() to run the PCA.
# Note there are 6 different Principal Components (PC) for the 6 different variables I selected.

pca_country_env <- env_country %>%
  select(-country) %>%
  scale() %>%
  prcomp()



# Look at the loadings for each variable

pca_country_env$rotation
  
```

### Part 1 PCA Biplot


```{r}

# Make a biplot to visualize the PCA results.
# Use autoplot() function to create a biplot, it is built to understand the PCA outcomes. 
# Set loadings = TRUE to display arrows that show loadings.
# Set loadings.label = TRUE to label the loading arrows with the variable names. 
# Used this website https://cran.r-project.org/web/packages/ggfortify/vignettes/plot_pca.html to figure out how to better customize the biplot.

autoplot(pca_country_env,
         data = env_country,
         loadings = TRUE,
         loadings.size = 1.5,
         loadings.label = TRUE,
         loadings.colour = "blue",
         loadings.label.colour = "red",
         loadings.label.size = 3) +
  theme_minimal() +
  labs(title = "PCA Biplot of International Environmental and Climatic Data ", caption = "Biplot displaying the results of a PCA on six different environmental variables")


```

###  Part 2 Summary Results

- As the axis shows PC1 accounts for ~38.68% variance and PC2 accounts for ~30.44%. The total variance accounted for by the first two PC is 69.12%. 

- Mean annual rainfall and tree canopy cover are strongly correlated since the arrows are at almost the same angle.

- Slope and elevation are also fairly strongly correlated.

- Wind, slope, and elevation are correlated but not very strongly.

### Part 3 Scree Plot

For fun make a screeplot to display the fraction of variance in the data explained by each of the different 6 PC.

```{r}
# Make a dataframe with the PCA results.
# Used this website https://datavizpyr.com/how-to-make-scree-plot-in-r-with-ggplot2/ as an example.

var_explained_df <- data.frame(PC= paste0("PC",1:6),
                               var_explained=(pca_country_env$sdev)^2/sum((pca_country_env$sdev)^2))
 
# Look at dataframe. 

head(var_explained_df)

```

```{r}
# Make a scree plot using the var_explained_df created above.
# Use both geom_point() and geom_line() to show both lines and points.
# Use geom_text() function to label the points on the graph with the variance explained values set to 3 significant figures.

var_explained_df %>%
  ggplot(aes(x=PC,y=var_explained, group=1))+
  geom_point(size=4)+
  geom_line()+
  labs(title="Scree plot: PCA on Scaled Environmental Data", x = "PC", y = "Variance Explained", caption = "Fraction of variance accounted for by each PC") +
  geom_text(aes(label = sprintf("%.3f", var_explained)), color = "blue", vjust = 1, hjust = 1.5, digits = 4)
```

The graph above is a scree line and point plot that displays the fraction of variance accounted for by each of the six individual PCs. The blue numbers indicate the fraction of variance out of 1 accounted for by that PC. PC 1 accounts for 38.7% of the variance and PC 2 accounts for 30.4% of the variance. Notice, these percentages are the same as the axis labels in the PCA biplot for PC 1 and PC 2. This scree plot allows a way to visualize the remaining PCs variance that do not show up on the biplot and to see how variance is distributed among the different PCs. 

